Multivariate Analysis

### 2021-02-04 multivariate.Rmd
new RF model with continuous variables only

---
title: "Multivariate Analysis"
author: "Scott Coffin"
date: "1/19/2021"
output: html_document
---
# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load packages
library(tidyverse) #General everything
library(RColorBrewer)
library(ggplot2) #General plotting
library(ggrepel) #For adding text labels that repel away from data points
library(calecopal) #Color palette
library(shiny) #Runs shiny
library(shinythemes) #Shiny theme for the page
library(shinyWidgets) #Widgets
library(scales) #SSD - Use the percent format
library(reshape2) #Overview tab - melts bars together
library(ssdtools) #SSD package
library(DT) #Build HTML data tables
#library(plotly) #Make plots interactive
library(viridis) #Colors
library(scales) #To use "percent" function
library(shinyjs) #Exploration tab - reset button
library(tigerstats) #turns things into percents
library(ggbeeswarm) #plot all points
library(fitdistrplus) #alt SSD 
library(RColorBrewer) #colors
library(pheatmap) #pretty heat maps
library(rpart)  #for trees
#library(rattle)    # Fancy tree plot This is a difficult library to install (https://gist.github.com/zhiyzuo/a489ffdcc5da87f28f8589a55aa206dd) 
library(rpart.plot)             # Enhanced tree plots
library(RColorBrewer)       # Color selection for fancy tree plot
library(party)                  # Alternative decision tree algorithm
# library(partykit)               # Convert rpart object to BinaryTree
library(pROC)   #for ROC curves
library(uwot) #umap
library(ISLR)  #for the Carseat Data
```
## Data import
```{r data import}
# Load finalized dataset.
aoc <- read_csv("AquaticOrganisms_Clean_final.csv", guess_max = 10000)

@@

  mutate(dose.particles.mL.master.converted.reported = factor(dose.particles.mL.master.converted.reported))

@@

# final cleanup and factoring  
aoc_z$Species <- as.factor(paste(aoc_z$genus,aoc_z$species)) #must make value 'Species" (uppercase)
aoc_z$Group <- as.factor(aoc_z$organism.group) #must make value "Group"
aoc_z$Group <- fct_explicit_na(aoc_z$Group) #makes sure that species get counted even if they're missing a group
```
```{r}
# subset data to selected variables

multiVar <- aoc_z %>% dplyr::select(#doi, size.category, 
                                    size_f,
                                    size.length.um.used.for.conversions, 
                                    shape, 
                                    polymer, 
                                    particle.volume.um3, 
                                    density.mg.um.3, 
                                    organism.group,
                                    environment, 
                                    bio.org, #biological level of organization
                                    #af.time, #assessment factor based on exposure time
                                    treatments, #number of doses (no including control)
                                    effect, #yes no
                                    exposure.duration.d, 
                                    exposure.route, #Factor
                                    organism.group, #factor
                                    media.temp, #numeric
                                    lvl1_f, #endpoints
                                    lvl2_f, #endpoints
                                   # lvl3, 
                                     dose.mg.L.master, 
                                    sex, #factor
                                    media.ph, #numeric
                                    media.sal.ppt, #numeric
                                    dose.particles.mL.master,
                                   effect.metric, #NOEC LOEC
                                    functional.group, #factor
                                    charge, #positive or negatibe
                                    zetapotential.mV, # numeric   
                                   max.size.ingest.mm) %>%  #max ingestible size
  filter(!size_f == "Not Reported")  #take out not reported

#recode variables
multiVar <- multiVar %>% mutate(effect_10 = case_when(
    effect == "Y" ~ 1,
    effect == "N" ~ 0
  ))# %>% 
  #mutate_all(is.character, ~as.factor())

@@

```{r}
require()
#convert to matrix and transpose
transposed_size <- as.data.frame(t(as.matrix(CompletenessSummary_size[2:25]))) %>% #1 is category, 2-6 are variables
  arrange('1nm < 100nm')
#reassign column names
colnames(transposed_size) <- c("1nm < 100nm", "100nm < 1µm", "1µm < 100µm", "100µm < 1mm", "1mm < 5mm")
#transposedTag
#format as matrix
MissingMatrix_size <- data.matrix(transposed_size)
#build heatmap
pheatmap(MissingMatrix_size,
                           main = "Data Completeness by Size Bin", #title
                           fontsize = 12,
                           cluster_rows = FALSE, cluster_cols = FALSE,#disable dendrograms
                           display_numbers = TRUE,
                           treeheight_row = 0, treeheight_col = 0, #keeps clustering after dropping dendrograms
                           col = rev(brewer.pal(n = 9, name = "PuBu"))) #blue color scheme with 9 colors)

@@

```{r}
require()
#convert to matrix and transpose
transposed_environment <- as.data.frame(t(as.matrix(CompletenessSummary_environment[2:25])))
#reassign column names
colnames(transposed_environment) <- c("Freshwater", "Marine", "Terrestrial")
#transposedTag
#format as matrix
MissingMatrix_environment <- data.matrix(transposed_environment)
#build heatmap
pheatmap(MissingMatrix_environment,
                           main = "Data Completeness by environment", #title
                           fontsize = 12,
                           cluster_rows = FALSE, cluster_cols = FALSE,#disable dendrograms
                           display_numbers = TRUE,
                           treeheight_row = 0, treeheight_col = 0, #keeps clustering after dropping dendrograms
                           col = rev(brewer.pal(n = 9, name = "PuBu"))) #blue color scheme with 9 colors)

@@

```{r}
#PCA
pca <- stats::prcomp(multiVar2[,-5], retx = TRUE, rank. = 2)
#build color pallete
my_colors = colorRampPalette(c("red", "yellow", "green"))(nrow(multiVar2))
#plot
embed_plot(pca$x, multiVar2$polymer, color_scheme = palette.colors(palette = "Okabe-Ito"), #turbo, #rainbow, #my_colors, 
           title = "Polymer PCA", alpha_scale = 0.5, equal_axes = TRUE)
```

```{r}
require(plotly)
#PCA for discrete variables
multiVar_discrete <- multiVar %>% 
  select(size.length.um.used.for.conversions, polymer, dose.mg.L.master, exposure.duration.d) %>% 
  drop_na %>% 
  mutate_if(~is.numeric(.) && (.) > 0, log10)

#buildPCA
pca_discrete <- stats::prcomp(multiVar_discrete[,-2], retx = TRUE, rank. = 2)

embed_plotly(pca_discrete$x, multiVar_discrete$polymer, color_scheme = palette.colors(palette = "Okabe-Ito"), 
           title = "Polymer PCA", alpha_scale = 0.5, equal_axes = TRUE,
           tooltip = paste("Polymer:", multiVar_discrete$polymer))
```

## Random Forest
### Recursive Partitioning And Regression Trees

@@

It is important to note that the algorithm works by making the best possible choice at each particular stage, without any consideration of whether those choices remain optimal in future stages. That is, the algorithm makes a locally optimal decision at each stage. It is thus quite possible that such a choice at one stage turns out to be sub-optimal in the overall scheme of things.  In other words,  the algorithm does not find a globally optimal tree.
```{r}
#trim data so effect is always known
multiVar_sub <- multiVar %>% 
  drop_na(effect_10)

@@

#### Pruning

@@

### CForest
```{r run model}
require(party)
crf <- cforest(as.factor(effect_10) ~ size.length.um.used.for.conversions + particle.volume.um3 + exposure.duration.d +
                 media.temp + dose.particles.mL.master,
               controls = cforest_control(ntree = 500,
                                          mincriterion = qnorm(0.8), 
                                          trace = TRUE), # adds project bar because it's very slow
               data = multiVar_sub)

@@

```{r}
crf.prob <- matrix(unlist(crfsrvy.prob), ncol=2, byrow=TRUE)
apply(crf.prob,2,mean)
#[1] 0.2524514 0.7475486
tab <- round(cbind(by(rpart.prob[,2], INDICES=t1$where, mean), by(crf.prob[,2], INDICES=t1$where, mean)), 4)
colnames(tab) <- c("rpart", "cforest")
kable(tab)
```
# Non-linear Dose Response
If we hold all paramaters constant except for dose, we should observe a non-linear dose-response behavior that can be best predicted by a 3- or 4-parameter Hill Equation. The most available information is likely for polystyrene spheres in crustaceans.There are 21 studies available.

>>>>

---
title: "Multivariate Analysis"
author: "Scott Coffin"
date: "1/19/2021"
output: html_document
---
# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load packages
library(tidyverse) #General everything
library(RColorBrewer)
library(ggplot2) #General plotting
library(ggrepel) #For adding text labels that repel away from data points
library(calecopal) #Color palette
library(shiny) #Runs shiny
library(shinythemes) #Shiny theme for the page
library(shinyWidgets) #Widgets
library(scales) #SSD - Use the percent format
library(reshape2) #Overview tab - melts bars together
library(ssdtools) #SSD package
library(DT) #Build HTML data tables
#library(plotly) #Make plots interactive
library(viridis) #Colors
library(scales) #To use "percent" function
library(shinyjs) #Exploration tab - reset button
library(tigerstats) #turns things into percents
library(ggbeeswarm) #plot all points
library(fitdistrplus) #alt SSD 
library(RColorBrewer) #colors
library(pheatmap) #pretty heat maps
library(rpart)  #for trees
#library(rattle)    # Fancy tree plot This is a difficult library to install (https://gist.github.com/zhiyzuo/a489ffdcc5da87f28f8589a55aa206dd) 
library(rpart.plot)             # Enhanced tree plots
library(RColorBrewer)       # Color selection for fancy tree plot
library(party)                  # Alternative decision tree algorithm
# library(partykit)               # Convert rpart object to BinaryTree
library(pROC)   #for ROC curves
library(uwot) #umap
library(ISLR)  #for the Carseat Data
library(quantregForest)
library(caret)
library(tidyverse)
library(tidymodels)
library(skimr)
library(sf)
library(ggspatial)
library(nhdplusTools)
library(patchwork)
library(Metrics)
library(gt)
```
## Data import
```{r data import}
# Load finalized dataset.
aoc <- read_csv("AquaticOrganisms_Clean_final.csv", guess_max = 10000)

#assign row ID
aoc <- rowid_to_column(aoc, "ID")
aoc <- aoc %>% 
  mutate(ID = paste0(ID, "ID"))

@@

  mutate(dose.particles.mL.master.converted.reported = factor(dose.particles.mL.master.converted.reported)) %>% 
  mutate(af.time_noNA = replace_na(af.time, "Unavailable")) %>% 
  mutate(acute.chronic_f = factor(case_when(af.time_noNA == 10 ~ "Acute",
                                            af.time_noNA == 1 ~ "Chronic",
                                            af.time_noNA == "Unavailable" ~ "Unavailable"))) %>%    #factorize assesment factor time into chronic/acute
  mutate(dose.mg.L.master.AF.noec = dose.mg.L.master * af.noec) %>% #converts ECxx, LC50, HONEC, LOEC, NOEC all to NOEC
  mutate(dose.particles.mL.master.AF.noec = dose.particles.mL.master * af.noec)  #converts ECxx, LC50, HONEC, LOEC, NOEC all to NOEC

@@

# final cleanup and factoring  
aoc_z$Species <- as.factor(paste(aoc_z$genus,aoc_z$species)) #must make value 'Species" (uppercase)
aoc_z$Group <- as.factor(aoc_z$organism.group) #must make value "Group"
aoc_z$Group <- fct_explicit_na(aoc_z$Group) #makes sure that species get counted even if they're missing a group
```
```{r}
# subset data to selected variables
multiVar <- aoc_z %>% dplyr::select(#doi, size.category, 
                                    ID,
                                    size_f,
                                    size.length.um.used.for.conversions, 
                                    shape, 
                                    polymer, 
                                    particle.volume.um3, 
                                    density.mg.um.3, 
                                    organism.group,
                                    environment, 
                                    bio.org, #biological level of organization
                                    #af.time, #assessment factor based on exposure time
                                    treatments, #number of doses (no including control)
                                    effect_f, #yes no
                                    exposure.duration.d, 
                                    exposure.route, #Factor
                                    organism.group, #factor
                                    media.temp, #numeric
                                    lvl1_f, #endpoints
                                    lvl2_f, #endpoints
                                   # lvl3, 
                                     dose.mg.L.master, 
                                    sex, #factor
                                    media.ph, #numeric
                                    media.sal.ppt, #numeric
                                    dose.particles.mL.master,
                                   effect.metric, #NOEC LOEC
                                    functional.group, #factor
                                    charge, #positive or negatibe
                                    zetapotential.mV, # numeric   
                                   max.size.ingest.mm,#max ingestible size
                                   acute.chronic_f,
                                   dose.mg.L.master.AF.noec,
                                   dose.particles.mL.master.AF.noec) %>%  
  filter(!size_f == "Not Reported") %>%   #take out not reported 
  mutate_if(is.character, as.factor)

# #recode variables
# multiVar <- multiVar %>% mutate(effect_10 = case_when(
#     effect_f == "Y" ~ 1,
#     effect_f == "N" ~ 0
#   ))# 
```

@@

```{r}
require()
#convert to matrix and transpose
transposed_size <- as.data.frame(t(as.matrix(CompletenessSummary_size[2:28]))) %>% #1 is category, 2-6 are variables
  arrange('1nm < 100nm')
#reassign column names
colnames(transposed_size) <- c("1nm < 100nm", "100nm < 1µm", "1µm < 100µm", "100µm < 1mm", "1mm < 5mm")
#transposedTag
#format as matrix
MissingMatrix_size <- data.matrix(transposed_size)
#build heatmap
pheatmap(MissingMatrix_size,
                           main = "Data Completeness by Size Bin", #title
                           fontsize = 12,
                           cluster_rows = FALSE, cluster_cols = FALSE,#disable dendrograms
                           display_numbers = TRUE,
                           treeheight_row = 0, treeheight_col = 0, #keeps clustering after dropping dendrograms
                           col = rev(brewer.pal(n = 9, name = "PuBu"))) #blue color scheme with 9 colors)

@@

```{r}
require()
#convert to matrix and transpose
transposed_environment <- as.data.frame(t(as.matrix(CompletenessSummary_environment[2:28])))
#reassign column names
colnames(transposed_environment) <- c("Freshwater", "Marine", "Terrestrial")
#transposedTag
#format as matrix
MissingMatrix_environment <- data.matrix(transposed_environment)
#build heatmap
pheatmap(MissingMatrix_environment,
                           main = "Data Completeness by environment", #title
                           fontsize = 12,
                           cluster_rows = FALSE, cluster_cols = FALSE,#disable dendrograms
                           display_numbers = TRUE,
                           treeheight_row = 0, treeheight_col = 0, #keeps clustering after dropping dendrograms
                           col = rev(brewer.pal(n = 9, name = "PuBu"))) #blue color scheme with 9 colors)

@@

```{r}
#PCA
pca <- stats::prcomp(multiVar2[,-4], retx = TRUE, rank. = 2)
#build color pallete
my_colors = colorRampPalette(c("red", "yellow", "green"))(nrow(multiVar2))
#plot
embed_plot(pca$x, multiVar2$polymer, color_scheme = palette.colors(palette = "Okabe-Ito"), #turbo, #rainbow, #my_colors, 
           title = "Polymer PCA", alpha_scale = 0.5, equal_axes = TRUE)
```

```{r}
require(plotly)
#PCA for discrete variables
multiVar_discrete <- multiVar %>% 
  select(size.length.um.used.for.conversions, polymer, dose.mg.L.master, exposure.duration.d, effect) %>% 
  drop_na %>% 
  mutate_if(~is.numeric(.) && (.) > 0, log10) %>% 
  mutate(effect_10 = case_when(
    effect == "Y" ~ 1,
    effect == "N" ~ 0
  )) %>% 
  select(-effect)

#buildPCA
pca_discrete <- stats::prcomp(multiVar_discrete[,-2], retx = TRUE, rank. = 5, scale. = TRUE)

embed_plot(pca_discrete$x, multiVar_discrete$polymer)

embed_plotly(pca_discrete$x, multiVar_discrete$polymer, color_scheme = palette.colors(palette = "Okabe-Ito"), 
           title = "Polymer PCA", alpha_scale = 0.5, equal_axes = TRUE,
           tooltip = paste("Polymer:", multiVar_discrete$polymer))

#embed_plotly also option
```

# Random Forest
## Recursive Partitioning And Regression Trees

@@

It is important to note that the algorithm works by making the best possible choice at each particular stage, without any consideration of whether those choices remain optimal in future stages. That is, the algorithm makes a locally optimal decision at each stage. It is thus quite possible that such a choice at one stage turns out to be sub-optimal in the overall scheme of things.  In other words,  the algorithm does not find a globally optimal tree.

The size of the data set usually imposes a hard limit on how many features may be considered due to Curse of Dimensionality, i.e. your data becomes sparser and sparser as you increase the number of features considered, which usually leads to overfitting. While there is no rule of thumb relating to how many features vs.  the number of observations you should use, I try to keep $e^Nf < No$ (Nf = number of features, No = number of observations) to minimise overfitting [this is not always possible and it does not ensure that we won’t overfit]. In this case, our training set has 3200, which suggests that using 8 features (e^8 ~ 2980) is not entirely absurd. Obviously, this depends on your data, so we will cover some further overfitting checks later on.
```{r}
#trim data so effect is always known
multiVar_sub <- multiVar %>% 
  drop_na(effect_10)

@@

### Pruning

@@

## CForest
```{r run model}
require(party)
crf <- cforest(as.factor(effect_10) ~ size.length.um.used.for.conversions + particle.volume.um3 + exposure.duration.d +
                 media.temp + dose.particles.mL.master,
               controls = cforest_control(ntree = 500,
                                          mincriterion = qnorm(0.8), 
                                          trace = TRUE), # adds project bar because it's very slow
               data = multiVar_sub)

@@

```{r}
crf.prob <- matrix(unlist(crfsrvy.prob), ncol=2, byrow=TRUE)
apply(crf.prob,2,mean)
#[1] 0.2524514 0.7475486
tab <- round(cbind(by(rpart.prob[,2], INDICES=t1$where, mean), by(crf.prob[,2], INDICES=t1$where, mean)), 4)
colnames(tab) <- c("rpart", "cforest")
kable(tab)
```
# Random Forest Package

## Preparation
```{r}
library(quantregForest)
library(caret)
library(tidyverse)
library(tidymodels)
library(skimr)
library(sf)
library(ggspatial)
library(nhdplusTools)
library(patchwork)
library(Metrics)
library(gt)
```

```{r}
multiVar <- multiVar %>% 
  mutate_if(is.character, as.factor) %>% 
  mutate(effect_10 = case_when( #convert ordinal to numeric
     effect_f == "Yes" ~ 1,
     effect_f == "No" ~ 0))

#choose relevant predictors and log-transform
multiVar_small_log <- multiVar %>% 
  dplyr::select(ID, size_f, size.length.um.used.for.conversions, shape, polymer, particle.volume.um3, density.mg.um.3, organism.group, bio.org, treatments, effect_f, exposure.duration.d, exposure.route, lvl1_f, dose.mg.L.master) %>% 
  mutate_if(~is.numeric(.) && (.) > 0, log10) %>% 
  drop_na() %>%  #drop missing
 mutate(effect_10 = case_when( #convert ordinal to numeric
     effect_f == "Yes" ~ 1,
     effect_f == "No" ~ 0
   ))# %>% 
 #  select(-effect_f)

#no log transform for comparison
multiVar_small <- multiVar %>% 
  dplyr::select(ID, size_f, size.length.um.used.for.conversions, shape, polymer, particle.volume.um3, density.mg.um.3, organism.group, bio.org, treatments, effect_f, #effect_10 , 
                exposure.duration.d, exposure.route, lvl1_f, dose.mg.L.master) %>% 
  drop_na() #%>%  #drop missing
# mutate(effect_10 = case_when( #convert ordinal to numeric
#     effect == "Y" ~ 1,
#     effect == "N" ~ 0
#   )) %>% 
  # select(-effect)

#ensure completeness
skim(multiVar_small)
```
## Training Data

```{r}
# Create calibration and validation splits with tidymodels initial_split() function.
set.seed(4)
multiVar_small_split <- multiVar_small %>%
  initial_split(prop = 0.75, strata = polymer) # splits data into training and testing set.
# default is 3/4ths split (but 75% training, 25% testing).
# Stratification (strata) = grouping training/testing sets by region, state, etc.
# Using the "strata" call ensures the number of data points in the training data is equivalent to the proportions in the original data set. (Strata below 10% of the total are pooled together.)

# Create a training data set with the training() function
# Pulls from training and testing sets created by initial_split()
multiVar_small_train <- training(multiVar_small_split)
multiVar_small_test <- testing(multiVar_small_split)
# Examine the environment to be sure # of observations looks like the 75/25 split. 3199:1066.

# Create a separate dataset of available IDs that were not used in the training dataset.
notTrain <- aoc %>% # all COMIDS from StreamCat data, sampled or not
  filter(!ID %in% aoc$ID) # Removing sites used to train the model. n = 140,097
```
## Kitchen Sink Model

```{r}
# Step Three - Kitchen Sink model -----------------------------------------

# Create finalized training dataset and include all possible variables. 
rf_dat <- multiVar_small_train %>%
  select(-ID)

# Random forest -- 
# a decision tree model, using predictors to answer dichotomous questions to create nested splits.
# no pruning happens - rather, multiple trees are built (the forest) and then you are looking for consensus across trees
# training data goes down the tree and ends up in a terminal node.
# if testing data goes down the same route, then this upholds our conclusions. Or, if it goes awry, this allows us to look for patterns in how it goes awry.

set.seed(2) # assures the data pulled is random, but sets it for the run below (makes outcome stable)
myrf <- randomForest(y = rf_dat$effect_f, # dependent variable
  x = rf_dat %>%
    select(-effect_f), # selecting all predictor variables
  importance = T, # how useful is a predictor in predicting values (nothing causal)
  proximity = T, 
  ntrees = 100) # 500 trees default.

myrf # examine the results.
```
~20% error rate. Let's compare this with the same model with log-transformed values.
```{r}
# Create calibration and validation splits with tidymodels initial_split() function.
set.seed(4)
multiVar_small_log_split <- multiVar_small_log %>%
  initial_split(prop = 0.75, strata = polymer) # splits data into training and testing set.

# Create a training data set with the training() function
# Pulls from training and testing sets created by initial_split()
multiVar_small_log_train <- training(multiVar_small_log_split)
multiVar_small_log_test <- testing(multiVar_small_log_split)

# Create finalized training dataset and include all possible variables. 
rf_dat_log <- multiVar_small_log_train %>%
  select(-ID, -effect_10)

set.seed(2) # assures the data pulled is random, but sets it for the run below (makes outcome stable)
myrf_log <- randomForest(y = rf_dat_log$effect_f, # dependent variable
  x = rf_dat_log %>%
    select(-effect_f), # selecting all predictor variables
  importance = T, # how useful is a predictor in predicting values (nothing causal)
  proximity = T, 
  ntrees = 100) # 500 trees default.

myrf_log # examine the results.
```
No performance enhancement with log-transformed values.

## Continuous Predictors
Repeat with continous variables whenever possible, and max of 8 predictors.
```{r}
skim(multiVar)
```
```{r}
acute <- multiVar %>% filter(acute.chronic_f == "Chronic") %>% 
  filter(!environment == "Terrestrial") %>% 
  filter(bio.org == "organism") %>% 
  filter(lvl1_f == "Fitness") %>% 
  filter(!exposure.route == "food") %>% 
  select(-c(sex, functional.group, charge, zetapotential.mV, max.size.ingest.mm, media.sal.ppt, media.ph, media.temp, treatments, acute.chronic_f, max.size.ingest.mm, effect.metric, lvl2_f, exposure.route, exposure.duration.d,bio.org, environment, lvl1_f, dose.particles.mL.master.AF.noec, dose.mg.L.master.AF.noec, size_f, density.mg.um.3, dose.particles.mL.master)) %>% 
  mutate_if(~is.numeric(.) && (.) > 0, log10) %>% 
  drop_na()
skim(acute)
```
With a complete dataset for just acute studies in aquatic organisms with aqueous route of exposure, we are left with 453 complete cases with 6 predictor variables, 1 response variable (effect y/n), and an ID. Let's determine if we have enough data for ML. 
```{r}
exp(1)^6
```
$e^6$ is less than n (453), so we may proceed.

```{r}
plot(myrf)
# model performance appears to improve most at ~75 trees
```

```{r}
varImpPlot(myrf)
# displays which variables are most important
# helps to winnow down list of predictors
# recommended to weigh left pane more
# right pane also shows how evenly things split based on the list of predictors
# values close to 0 can be dropped, but don't have to be
```
Dose, organism group, exposure duration are important.

```{r}
importance <- myrf$importance
View(importance)
# displays the data plotted in the plot above
```

```{r}
# predict()
# returns out of bag predictions for training data
# in the bag: every time a tree is built, it uses ~80% of the original 75% we set aside from the original dataset used to create a tree to assure random data selection
# out of bag: looking at the remaining 20% of the training data to predict, when you want to know what your model does at the training location sites

# # Predict CRAM scores state-wide for all COMIDs.
# notTrain_prediction <- notTrain %>% # taking all available IDs, that haven't been used in training
#   na.omit() %>% # remove NAs
#   mutate(effect_predicted = predict(myrf, newdata = notTrain %>% na.omit())) # using developed model (myrf), inputting predictor variables (notTrain - which contains IDs and associated  data) to predict output/dependent variable (effect_predicted a.k.a. effect).

# rePredict CRAM scores for training data.
multiVar_small_train$effect_predicted <- predict(myrf) # Add column of predicted CRAM values to training dataset.

# Creates new dataset of bound rows for both ...
# effect_predictions <- bind_rows(nottrain_prediction %>%
#                             mutate(Set = "Non-training"), # statewide COMIDs (not used for training)
#                             multiVar_smal_train %>%
#                             mutate(Set = "Training")) # COMIDS from training dataset
# This creates the dataset that will be plotted to create a state-wide plot of predicted CRAM scores.

# Plot the data.
multiVar_small_train %>% 
  mutate(effect_10 = case_when( #convert ordinal to numeric
     effect_predicted == "Yes" ~ 1,
     effect_predicted == "No" ~ 0
   )) %>% 
  ggplot(aes(x = log10(dose.mg.L.master), y = effect_10)) +
  geom_point(alpha = 0.7) +#color = "black") +
  labs(x = "Density",
    y = "Predicted CRAM Score") +
  theme_dark() #+
  #facet_wrap(.~polymer)
```
### Predictor Selector
```{r}
# Using caret to select the best predictors
# What are the parameters you want to use to run recursive feature elimination (rfe)?
my_ctrl <- rfeControl(functions = rfFuncs,
                      method = "cv",
                      verbose = FALSE,
                      returnResamp = "all")

# rfe = recursive feature elimination
# THIS STEP TAKES FOR-EV-ER!!!
set.seed(22)
my_rfe <- rfe(y = rf_dat$effect_f, # set dependent variable
              x = rf_dat %>% select(-effect_f), # set predictor variables
              size = c(1:2, 4, 6, 8, 10, 13), # sets how many variables are in the overall model
              # I have 13 total possible variables, so I've chosen increments of 3 to look at.
              rfeControl = my_ctrl) # pull in control from above

# can you make your model even simpler?
# the following will pick a model with the smallest number of predictor variables based on the tolerance ("tol") that you specify (how much less than the best are you willing to tolerate?)
my_size <- pickSizeTolerance(my_rfe$results, metric = "Accuracy", tol = 1, maximize = F)
# higher tol (~10) gives you less variables
# lower tol (~1) gives you more variables - "I'd like the simplest model within 1% of the best model."
pickVars(my_rfe$variables, size = my_size)
```

```{r}
# Predict scores using the above 2 variables:

# Create re-finalized training dataset and include all possible variables. 
rf_dat2 <- multiVar_small_train %>%  select(dose.mg.L.master, organism.group, effect_f)

set.seed(4) # assures the data pulled is random, but sets it for the run below (makes outcome stable)
myrf2 <- randomForest(y = rf_dat2$effect_f, # dependent variable
  x = rf_dat2 %>%
    select(-effect_f),
  importance = T, 
  proximity = T, 
  ntrees = 500)

myrf2 # examine the results. 
```

```{r}
importance <- as.data.frame(as.table(myrf$importance))

# Nicer ggplot variable importance plot.
vip_plot_a <- importance %>%
 filter(Var2 == "MeanDecreaseAccuracy") %>%
  mutate(Var1 = factor(Var1)) %>%
  mutate(Var1_f = fct_reorder(Var1, Freq)) %>%
  ggplot(aes(x = Freq, y = Var1_f)) +
  geom_point(size = 3, alpha = 0.75, color = "black") +
  labs(x = "Importance",
    y = "Variables") +
  theme_bw()

# vip_plot_b <- importance %>%
#   filter(Var2 == "MeanDecreaseGini") %>%
#   mutate(Var1 = factor(Var1)) %>%
#   mutate(Var1_f = fct_reorder(Var1, Freq)) %>%
#   ggplot(aes(x = Freq, y = Var1_f)) +
#   geom_point(size = 3, alpha = 0.75, color = "black") +
#   labs(x = "Node Purity",
#     y = "Variables") +
#   theme_bw()

vip_plot <- vip_plot_a #+ vip_plot_b

vip_plot
# ggsave("cram_vip_plot.png",
#      path = "/Users/heilil/Desktop/R_figures",
#      width = 25,
#      height = 10,
#      units = "cm"
#    )
```
GINI importance measures the average gain of purity by splits of a given variable. If the variable is useful, it tends to split mixed labeled nodes into pure single class nodes. Splitting by a permuted variables tend neither to increase nor decrease node purities. Permuting a useful variable, tend to give relatively large decrease in mean gini-gain. GINI importance is closely related to the local decision function, that random forest uses to select the best available split. Therefore, it does not take much extra time to compute. On the other hand, mean gini-gain in local splits, is not necessarily what is most useful to measure, in contrary to change of overall model performance. Gini importance is overall inferior to (permutation based) variable importance as it is relatively more biased, more unstable and tend to answer a more indirect question.

### Quantile Regression model
```{r}
# Quantile random forest regression mode, instead of looking at the mode of trees, can compare to 10th, 50th, 90th percentiles etc.

# Need to make a new dataset taking the above results of pickVars into account.
# Create finalized training dataset and include all possible variables. 
qrf_dat <- multiVar_small_train  %>%
  select(-effect_predicted, -ID)
#   select(asci, RdCrsWs, PctAgWs, PctUrbWsRp100, PctOpWsRp100, PctOpWs, DamDensWs, RdDensWs, NABD_DensWs, PctUrbWs, PctUrbCatRp100, RdDensWsRp100, PctOpCat, PctUrbCat, RdDensCat, CBNFWs, PctOpCatRp100, PctAgWsRp100, TRIDensWs, AgKffactWs, FertWs)

 set.seed(20)
 myqrf <- quantregForest(y = qrf_dat$effect_f, # dependent variable
              x = qrf_dat %>%
                  select(-effect_f),
              importance = T,
              proximity = T,
              keep.inbag=T,
              ntrees = 500)

#predict(myqrf) # automatically presents 10th %tile, median, and 90th %tile
```

```{r}
#predict(myqrf, what=c(0.2, 0.3, 0.999)) # to print specific quantiles
```

```{r}
plot(myqrf) # plots the results.
```
# Again appears to improve after ~100 trees.

### Model Validation
#### Categorical Response

#### Continuous Response
```{r}
# Compare predicted vs. actual results, including by size.
# Adding lines of slope=1 and linear models to each plot.
val1 <- multiVar_small_train %>% 
  ggplot(aes(x = effect_predicted, y = effect_f)) +
  geom_point(color = "#2A3927", alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "#2A3927") +
  labs(x = "Predicted Effect",
    y = "Effect measured",
    title = "Training Data\nn=613") +
  geom_abline(intercept = 0, slope = 1) +
  theme_bw()

val1

lm1 <- lm(cram~cram_predicted, data = mydf2_train2)
summary(lm1)

val2 <- ggplot(mydf2_test2, aes(x = cram_predicted, y = cram)) +
  geom_point(color = "#3793EC", alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "#3793EC") +
  scale_x_continuous(breaks = c(0.5, 0.7, 0.9)) +
  labs(x = "CRAM predicted",
    y = "CRAM measured",
    title = "Testing Data\nn=202") +
  geom_abline(intercept = 0, slope = 1) +
  theme_bw()

val2

lm2 <- lm(cram~cram_predicted, data = mydf2_test2)
summary(lm2)

# Create the full testing + training dataset to plot together.
mydf2_test2$set <- "Testing"
mydf2_train2$set <- "Training"
full_train_test <- bind_rows(mydf2_test2, mydf2_train2) %>%
  mutate(set_f = factor(set, levels = c("Training", "Testing")))

val3 <- ggplot(full_train_test, aes(x = cram_predicted, y = cram, color = set_f)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(name = "Set", values = c("#2A3927", "#3793EC"), drop = FALSE) +
  labs(x = "CRAM predicted",
    y = "CRAM measured",
    title = "All Data\nn=815") +
  geom_abline(intercept = 0, slope = 1, color = "black") +
  facet_wrap(~PSA6) +
  theme_bw()

val3

val_fig <- (val1 + val2) /
  (val3)

val_fig + plot_annotation(
  title = 'CRAM Random Forest Results',
  subtitle = 'All modeling performed using StreamCAT datasets.',
  caption = 'Linear models are colored according to dataset. Lines of slope = 1 are denoted in black.'
)

# Save figure.
# ggsave("cram_rfmodel_validation.png",
#      path = "/Users/heilil/Desktop/R_figures",
#      width = 35,
#      height = 25,
#      units = "cm"
#    )

lm3 <- lm(cram~cram_predicted, 
  data = full_train_test %>%
    filter(PSA6 == "Central_Valley") %>%
    filter(set_f == "Training"))

lm4 <- lm(cram~cram_predicted, 
  data = full_train_test %>%
    filter(PSA6 == "Central_Valley") %>%
    filter(set_f == "Testing"))

lm5 <- lm(cram~cram_predicted, 
  data = full_train_test %>%
    filter(PSA6 == "Chaparral") %>%
    filter(set_f == "Training"))

lm6 <- lm(cram~cram_predicted, 
  data = full_train_test %>%
    filter(PSA6 == "Chaparral") %>%
    filter(set_f == "Testing"))

lm7 <- lm(cram~cram_predicted, 
  data = full_train_test %>%
    filter(PSA6 == "Deserts_Modoc") %>%
    filter(set_f == "Training"))

lm8 <- lm(cram~cram_predicted, 
  data = full_train_test %>%
    filter(PSA6 == "Deserts_Modoc") %>%
    filter(set_f == "Testing"))

lm9 <- lm(cram~cram_predicted, 
  data = full_train_test %>%
    filter(PSA6 == "North_Coast") %>%
    filter(set_f == "Training"))

lm10 <- lm(cram~cram_predicted, 
  data = full_train_test %>%
    filter(PSA6 == "North_Coast") %>%
    filter(set_f == "Testing"))

lm11 <- lm(cram~cram_predicted, 
  data = full_train_test %>%
    filter(PSA6 == "Sierra") %>%
    filter(set_f == "Training"))

lm12 <- lm(cram~cram_predicted, 
  data = full_train_test %>%
    filter(PSA6 == "Sierra") %>%
    filter(set_f == "Testing"))

lm13 <- lm(cram~cram_predicted, 
  data = full_train_test %>%
    filter(PSA6 == "South_Coast") %>%
    filter(set_f == "Training"))

lm14 <- lm(cram~cram_predicted, 
  data = full_train_test %>%
    filter(PSA6 == "South_Coast") %>%
    filter(set_f == "Testing"))

# Chose not to compute confusion matrix / accuracy score since this is more applicable to categorical ouputs from random forest models -
# Instead, calculated Root Mean Squared Error (RMSE) of both training and test datasets.
# If test RMSE values are much greater than training, then possible the model has been over fit.

predtest <- predict(myrf2, mydf2_test2)
rmse(mydf2_test2$cram,predtest)
# 9.03
